Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-10-01 paper/intro.tex
try to fix pipeline

\textcolor{red}{??? The following two paragraphs need to rewritten a bit at a later stage. It needs to be clearer what the approaches proposed in our paper are really about and the novelty wrt existing work has to come out more clearly and concretely. ???}
\color{black}
Instead of discretizing $[a, b]$ into $t_1, \ldots, t_m$, approximating $\Trace(\mtx{B}(t_i))$ with $b(t_i)$, and analyzing the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way is to quantify the error with $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate in each $t \in [a, b]$. This approach has already been used in the analysis of low-rank approximations of parameter-dependent matrices \cite{kressner-2024-randomized-lowrank}.

\paragraph{New contributions.} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}. Interestingly, we can reuse the same randomization for computing the estimate at ea<ch value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations. We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}. In particular, we use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude. Based on the analysis of the randomized trace estimators for parameter-dependent matrices, we provide bounds on the approximation error for our method.

>>>>

\textcolor{red}{??? The following two paragraphs need to rewritten a bit at a later stage. It needs to be clearer what the approaches proposed in our paper are really about and the novelty wrt existing work has to come out more clearly and concretely. ??? Mention \cite{park-2024-lowrank-approximation} ?.}
\color{black}
Instead of discretizing $[a, b]$ into $t_1, \ldots, t_m$, approximating $\Trace(\mtx{B}(t_i))$ with $b(t_i)$, and analyzing the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way is to quantify the error with $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate in each $t \in [a, b]$. This approach has already been used in the analysis of low-rank approximations to parameter-dependent matrices \cite{kressner-2024-randomized-lowrank}.

\paragraph{New contributions.} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}. Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations. We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}. In particular, we use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude. Based on the analysis of the randomized trace estimators for parameter-dependent matrices, we provide bounds on the approximation error for our method.
